# Simulation and Randomization {#ch4}

```{r, include = F}
rm(list = ls(all = T))
knitr::opts_chunk$set(fig.align = "center")
```

## Chapter Overview{-}

Simulation modeling is one of the primary reasons to move away from spreadsheet-type programs (like Microsoft Excel) and into a program like R. R allows you to replicate the same (possibly complex and detailed) calculations over and over with slightly different random values. You can then summarize and plot the results of these replicated calculations all within the same program. Analyses of this type are **Monte Carlo methods**: they randomly sample from a set of quanities for the purpose of generating and summarizing a distribution of some statistic related to the sampled quantities. If this concept is confusing, hopefully this chapter will clarify. 

In this chapter, you will learn the basic skills needed for simulation (i.e., Monte Carlo) modeling in R including:

*  creation of random deviates
*  `for` loops
*  more advanced function writing
*  summarization of many values from a distribution

**IMPORTANT NOTE**: If you did not attend the sessions corresponding to Chapters \@ref(ch1) or \@ref(ch2), you are recommended to walk through the material found in those chapters before proceeding to this material. Additionally, you will find the material in Section \@ref(dists) helpful for this chapter. Remember that if you are confused about a topic, you can use **CTRL + F** to find previous cases where that topic has been discussed in this document.

## Before You Begin {-}

You should create a new directory and R script for your work in this Chapter. Create a new R script called `Ch4.R` and save it in the directory `C:/Users/YOU/Documents/R-Workshop/Chapter4`. Set your working directory to that location. Revisit the material in Sections \@ref(scripts) and \@ref(working-dir) for more details on these steps.

## Layout of this chapter

This chapter is divided into two main sections:

*  **Required material** (Sections \@ref(randomness) and \@ref(mc-summaries)) which is necessary to understand the examples in this chapter and the subsequent chapters

*  **Example Cases** (Sections \@ref(sim-examples) and \@ref(resample-examples)) which apply the skills learned in the required material. In the workshop sessions, you will walkthrough 2-3 of these example cases at the choice of the group of the participants. If you are interested in simulation modeling, you are suggested to work through all of the example cases, as slightly different tricks will be shown in the different examples. 

## Introducing Randomness {#randomness}

A critical part of simulation modeling is the use of random processes. A **random process** is one that generates a different outcome according to some rules each time it is executed. They are tightly linked to the concept of **uncertainty**: you are unsure about the outcome the next time the process is executed. There are two basic ways to introduce randomness in R: **random deviates** and **resampling**.

### Random deviates

In Section \@ref(dists), you learned about using probability distributions in R. One of the uses was the `r-` family of distribution functions. These functions create random numbers following a random process specified by a probability distribution. 

Consider animal survival as an example. At the end of each year, each individual alive at the start can either live or die. There are two outcomes here, and suppose each animal has an 80% chance of surviving. The number of individuals that survive is the result of a **binomial random process** in which there were $n$ individuals alive at the start of this year and $p$ is the probability that any one individual survives to the next year. You can execute one binomial random process where $p = 0.8$ and $n = 100$ like this:

```{r}
rbinom(n = 1, size = 100, prob = 0.8)
```

The result you get will almost certainly be different from the one printed here. That is the random component.

You can execute many such binomial processes by changing the `n` argument. Plot the distribution of expected surviving individuals:

```{r}
survivors = rbinom(1000, 100, 0.8)
hist(survivors, col = "skyblue")
```

Another random process is the **lognormal process**: it generates random numbers such that the log of the values are normally-distributed with mean equal to `logmean` and standard deviation equal to `logsd`:

```{r}
hist(rlnorm(1000, 0, 0.1), col = "skyblue")
```

There are many random processes you can use in R. Checkout Table \@ref(tab:dist-table) for more examples as well as the help files for each individual function for more details.

### Resampling

Using random deviates works great for creating new random numbers, but what if you already have a set of numbers that you wish to introduce randomness to? For this, you can use **resampling techniques**. In R, the `sample` function is used to sample `size` elements from the vector `x`:

```{r}
sample(x = 1:10, size = 5)
```

You can sample with replacement (where it is possible to sample the same element two or more times):

```{r}
sample(x = c("a", "b", "c"), size = 10, replace = T)
```

You can set probabilities on the sampling of different elements^[If `prob` doesn't sum to 1, then it will be rescaled: `prob = prob/sum(prob)`]:

```{r}
sample(x = c("live", "die"), size = 10, replace = T,
       prob = c(0.8, 0.2))
```

Notice that this is the same as the binomial random process above, but with only 10 trials and the printing of the outcomes rather than the number of successes.

## Reproducing randomness 

For reproducibility purposes, you may wish to get the same exact random numbers each time you run your script. To do this, you need to set the **random seed**, which is the starting point of the random number generator your computer uses. If you run these two lines of code, you should get the same result as printed here:

```{r}
set.seed(1234)
rnorm(1)
```

## Replication

To use Monte Carlo methods, you need to be able to replicate some random process many times. There are two main ways this is commonly done: either with`replicate` or with `for` loops.

### `replicate`

The `replicate` function executes some expression many times and returns the output from each execution. Say we have a vector `x`, which represents 30 observations of fish length (mm):

```{r}
x = rnorm(30, 500, 30)
```

We wish to build the sampling distribution of the mean length "by hand". We can sample randomly from it, calculate the mean, then repeat this process many times:

```{r}
means = replicate(n = 1000, expr = {
  x_i = sample(x, length(x), replace = T)
  mean(x_i)
})
```

If we take `mean(means)` and `sd(means)`, that should be very similar to `mean(x)` and `se(x)`. Create the `se` function (also shown in Section \@ref(error-bars)) and prove this to yourself:

```{r}
se = function(x) sd(x)/sqrt(length(x))
mean(means); mean(x)
sd(means); se(x)
```

### The `for` loop {#for-loops}

In programming, a *loop* is a command that does something over and over until it reaches some point that you specify. R has a few types of loops: `repeat`, `while`, and `for`, to name a few. `for` loops are among the most common in simulation modeling. A `for` loop repeats some action for however many times you tell it for each value in some vector. The syntax is:

```{r eval = F}
for (var in seq) {
  expression(var)
}
```

The loop calculates the expression for values of `var` for each element in the vector `seq`. For example:

```{r}
for (i in 1:5) {
  print(i^2)
}
```

The `print` command will be executed 5 times: once for each value of `i`. It is the same as:

```{r, eval = F}
i = 1; print(i); i = 2; print(i); i = 3; print(i); i = 4; print(i); i = 5; print(i)
```

If you remove the `print()` function, see what happens:

```{r}
for (i in 1:5) {
  i^2
}
```

Nothing is printed to the console. R did the calculation, but did not show you or store it. Often, you'll need to store the results of the calculation in a **container object**:

```{r}
results = numeric(5)
```

This makes an empty numeric vector of length 5 that are all 0's. You can store the output of your loop calculations in `results`:

```{r}
for (i in 1:5) {
  results[i]=i^2
}
results
```

When `i^2` is calculated, it will be placed in the element `results[i]`. This was a trivial example, because you should do things like this using R's vectorized calculation framework: `(1:5)^2` (see Section \@ref(vector-math)). 

However, there are times where it is advantageous to use a loop. Particularly in cases where:

1.  the calculations in one element are determined from the value in previous elements, such as in time series models
2.  the calculations have multiple steps 
3.  you wish to store multiple results
4.  you wish to track the progress of your calculations

As an illustration for item (1) above, build a (very) basic population model. At the start of the first year, the population abundance is 1000 individuals and grows by an average factor of 1.1 per year (reproduction and death processes result in a growth rate of 10%) before harvest. The growth rate varies randomly, however. Each year, the 1.1 growth factor has variability introduced by small changes in survival and reproductive process. Model these variations as lognormal random variables. After production, 8% of the population is harvested. Simulate and plot the abundance at the end of the year for 100 years:

```{r}
nt = 100        # number of years
N = NULL       # container for abundance
N[1] = 1000    # first end-of-year abundance

for (t in 2:nt) {
  # N this year is N last year * growth *
    # randomness * fraction that survive harvest
  N[t] = (N[t-1] * 1.1 * rlnorm(1, 0, 0.1)) * (1 - 0.08)
}

# plot
plot(N, type = "l", pch = 15, xlab = "Year", ylab = "Abundance")
```

Examples of the other three utilities are shown in the example cases.

## Function Writing {#adv-funcs}
In Monte Carlo analyses, it is often useful to wrap code into functions. This makes them easy to be replicated and have the settings adjusted. As an example, turn the population model shown above into a function:

```{r}
pop_sim = function(nt, grow, sd_grow, U, plot = F) {
  N = NULL 
  N[1] = 1000

  for (t in 2:nt) {
    N[t] = (N[t-1] * grow * rlnorm(1, 0, sd_grow)) * (1 - U)
  }
  
  if (plot) {
    plot(N, type = "l", pch = 15, xlab = "Year", ylab = "Abundance")
  }
  
  N
}
```

This function takes five inputs: 

*  `nt`: the number of years,
*  `grow`: the population growth rate,
*  `sd_grow`: the amount of annual variability in the growth rate
*  `U`: the annual exploitation rate
*  `plot`: whether you wish to have a plot created. It has a default setting of `FALSE`: if you don't specify `plot = T` when you call `pop_sim`, you won't see a plot made.

It returns one output: the vector of population abundance.

Use your function once using the same settings as before:

```{r}
pop_sim(100, 1.1, 0.1, 0.08, T)
```

Now, you wish to replicate exectuing this function 1000 times. Use the `replicate` function to do this:

```{r}
out = replicate(n = 1000, expr = pop_sim(100, 1.1, 0.1, 0.08, F))
```

If you do `dim(out)`, you'll see that rows are stored as years (there are `r nrow(out)` of them) and columns are stored as replicates (there are `r ncol(out)` of them). Notice how wrapping the code in the function made the `replicate` call easy. 

**Here are some advantages of wrapping code like this into a function**:

*  If you do the same task over and over, you don't need to type all of the code to perform the task, just the function call. 
*  If you need to change the way the function behaves (mechanically in the function body), you only need to change it one place: in the function definition.
*  You can easily change the settings of the code (e.g., whether you want to see the plot) in one place
*  Function writing **can** lead to shorter scripts
*  Function writing **can** lead to more readable code (if people know what your functions do)

## Summarization {#mc-summaries}

After replicating a calculation many times, you will need to summarize the results. Here are several examples using the `out` matrix.

### Central Tendency

You can calculate the mean abundance each year across your iterations using the `apply` function (Section \@ref(data-summaries)):

```{r}
N_mean = apply(out, 1, mean)
N_mean[1:10]
```

You could do the same thing using `median` rather than `mean`. Mode is more difficult to calculate in R, if you need to get the mode, try to Google it^[Google is an R programmer's best friend. There is a massive online community for R, and if you have a question on something, it has almost certainly been asked somewhere on the web.].

### Variability

One of the primary reasons to conduct a Monte Carlo analysis is to obtain estimates of variability. You can summarize the variability easily using the `quantile` function:

```{r}
N_quants = apply(out, 1, function(x) quantile(x, c(0.1, 0.9)))

plot(N_mean, type = "l", ylim = c(0, 10000))
lines(N_quants[1,], lty = 2)
lines(N_quants[2,], lty = 2)
```

Notice how a user-defined function was passed to `apply`. The range within the two dashed lines represents the range that encompassed the central 80% of the random abundances each year.

### Frequencies

Often you will want to count how many times something happened. In some cases, the fraction of times something happened can be interpretted as a probability.

The `table` function is very useful for counting occurences of events. Suppose you are interested in how many of your iterations resulted in fewer than 1000 individuals at year 10:

```{r}
out10 = ifelse(out[10,] < 1000, "less10", "greater10")
table(out10)
```

Suppose you are also interested in how many of your iterations resulted in fewer than 1100 individuals at year 20:

```{r}
out20 = ifelse(out[20,] < 1100, "less20", "greater20")
table(out20)
```

Now suppose you are interested in how these two metrics are related:

```{r}
table(out10, out20)
```

As an example in iterpretting this output, most often populations that were greater than 1000 at year 10 were also greater than 1100 at year 20. If a population was less than 1000 at year 10, it was more likely to be less than 1100 at year 20 than to be greater than it.

You can turn these into probabilities (if you believe your model represents reality) by dividing each cell by the total number of iterations:

```{r}
round(table(out10, out20)/1000, 2)
```

## Simulation-Based Examples {#sim-examples}

### Test `rnorm`

In this example, you will verify that the function `rnorm` works the same way that `qnorm` and `pnorm` work. Hopefully it will also reinforce the way the random, quantile, and cumulative distribution functions work in R. If you're feeling advanced, try adjusting this example for another distribution like the Poisson.

First, specify the mean and standard deviation for this example:

```{r}
mu = 500; sig = 30
```

Now make up `n` (any number of your choosing, something greater than 10) random deviates from this normal distribution:

```{r}
random = rnorm(100, mu, sig)
```

Test the quantiles (obtain the values that `p` * 100% of the quantities fall below, both for random numbers and from the `qnorm` function):

```{r}
p = seq(0.01, 0.99, 0.01)

random_q = quantile(random, p)
normal_q = qnorm(p, mu, sig)

plot(normal_q ~ random_q); abline(c(0,1))
```

The fact that all the quantiles fall around the 1:1 line sugggests the `n` random samples samples are indeed from a normal distribution. Any variabilities you see are due to sampling errors. If you increase `n` to `n = 1e6` (one million), you'll see no deviations.  This is called a **q-q plot**, and is frequently used to assess the fit of data to a distribution.

Now test the values in their agreement with the `pnorm` function. Plot the cumulative density functions for the truely normal curve and that approximated by the random deviates:

```{r}
q = seq(400, 600, 10)

random_cdf = ecdf(random)
random_p = random_cdf(q)
normal_p = pnorm(q, mu, sig)

plot(normal_p ~ q, type = "l", col = "blue")
points(random_p ~ q, col = "red")
```

The `ecdf` function obtains the empirical cumulative density function (which is just `pnorm` for a sample). It allows you to plug in any random variable and obtain the probability of having one less than it.

### Stochastic Power Analysis {#power-ex}

A power analysis is one where the analyst wishes to determine how much power they will have to detect an effect. Having high power means ensuring you do not falsely reject a true hypothesis (e.g., claiming that there is not effect based on a p-value greater than 0.05).

You can conduct a power analysis using stochastic simulation. Here, you will write a power analysis to determine how likely are you to be able to correctly identify what you deem to be a biologically-meaningful difference in survival between two tagging procedures. 

You know one tagging procedure has approximately a 10% mortality rate (10% of tagged fish die within the first 12 hours as result of the tagging process). Another, cheaper and less labor intensive method has been proposed but before implementing it, your agency wishes to determine if it will have a meaningful impact on the reliability of the study. You and your colleagues determine that if the mortality rate reaches 25%, then gains in time and cost efficiency would be offset by needing to tag more fish (because more will die). You have decided to perform a small-scale study to determine if the new method affects mortality enough to result in 25% or more mortality. The study will tag `n` individuals using each method (new and old) and track the fraction that survived after 12 hours. Before performing the study however, you deem it important to determine how large `n` needs to be to answer this question. You decide to use a stochastic power analysis based on what you've learned in this book to help your research group. The small-scale study can tag a total of at most 100 fish with the currently available resources. Could you tag fewer than 100 total individuals and still have a high probability of correctly identifying an effect of this size?

The stochastic power analysis approach works like this (this is called **psuedocode**):

1.  Simulate data under the reality that the difference is real with `n` observations per treatment, where `n < 30/2`. 
2.  Fit the model that will be used when the real data are collected.
3.  Determine if an effect at least as large as the effect of interest was detected with a significant p-value
4.  Replicate steps 1 - 3 many times. Determine what fraction of times the effect was correctly identified. 
5.  Replicate step 4 while varying `n` over the interval from 5 to 15.

Step 2 will require fitting a generalized linear model, for a review revisit Section \@ref(glms) (specifically Section \@ref(logis-regression) on logistic regression).

First, create a function that will generate data, fit the model, and determine if the p-value is significant (steps 1-3 above):

```{r}
sim_fit = function(n, p_old = 0.10, p_new = 0.25) {
  
  ### step 1: create the data ###
  # generate random response data
  dead_old = rbinom(n, size = 1, prob = p_old)
  dead_new = rbinom(n, size = 1, prob = p_new)
  
  # create the predictor variable
  method = rep(c("old", "new"), each = n)
  
  # create a data.frame to pass to glm
  df = data.frame(dead = c(dead_old, dead_new), method = method)
  
  # relevel so old is the reference
  df$method = relevel(df$method, ref = "old")
  
  ### step 2: fit the model ###
  fit = glm(dead ~ method, data = df, family = binomial)
  
  ### step 3: determine if a sig. p-value was found ###
  # extract the p-value
  pval = summary(fit)$coef[2,4]
  
  # determine if it was found to be significant
  pval < 0.05
}
```

Next, for steps 4 and 5, set up a **nested `for` loop**. This will have two loops: one that loops over sample sizes for step 5 and one that loops over replicates of each sample size (step 4):

```{r}
I = 10  # the number of replicates at each sample size

n_try = seq(10, 50, 10)  # the test sample sizes
N = length(n_try)      # count them

# container: 
out = matrix(NA, I, N) # matrix with I rows and N columns

for (n in 1:N) {
  for (i in 1:I) {
    out[i,n] = sim_fit(n = n_try[n])
  }
}
```

You now have a matrix of `TRUE` and `FALSE` elements that indicates whether a significant difference was found at the $\alpha = 0.05$ level if the effect was truely as large as you care about. You can obtain the proportion of all the replicates at each sample size that resulted in a significant difference using the `mean` function with `apply`:

```{r}
plot(apply(out, 2, mean) ~ n_try, type = "l",
     xlab = "Tagged Fish per Treatment",
     ylab = "Probability of Finding Effect (Power)")
```

Even if you tagged `r n_try[N] * 2` fish total, you would have a `r round(mean(out[,N]),2) * 100`% chance of saying the effect (which truely is there!) is present under the null hypothesis testing framework.

Suppose you and your colleagues aren't relying on p-values in this case, and are purely interested in how precisely the **effect size** would be estimated. Adapt your function to determine how frequently it is you would be able to estimate the true mortality probability of the new method within +/- 5% based on the point estimate only (the true estimate must be between 0.2 and 0.3 for a successful study). Change your function to calculate this additional metric and re-run the analysis:

```{r}
sim_fit = function(n, p_old = 0.10, p_new = 0.25) {
  
  # create the data
  dead_old = rbinom(n, size = 1, prob = p_old)
  dead_new = rbinom(n, size = 1, prob = p_new)
  
  # create the predictor variable
  method = rep(c("old", "new"), each = n)
  
  # create a data.frame to pass to glm
  df = data.frame(dead = c(dead_old, dead_new), method = method)
  
  # relevel so old is the reference
  df$method = relevel(df$method, ref = "old")
  
  # fit the model
  fit = glm(dead ~ method, data = df, family = binomial)
  
    ### step 3: determine if a sig. p-value was found ###
  # extract the p-value
  pval = summary(fit)$coef[2,4]
  
  # determine if it was found to be significant
  sig_pval = pval < 0.05
  
  # obtain the estimated mortality rate for the new method
  p_new_est = predict(fit, data.frame(method = c("new")),
                      type = "response")
  
  # determine if it is +/- 5% from the true value
  prc_est = p_new_est >= (p_new - 0.05) & p_new_est <= (p_new + 0.05)
  
  # return a vector with these two elements
  c(sig_pval = sig_pval, prc_est = unname(prc_est))
}

# run the analysis
I = 10  # the number of replicates at each sample size

n_try = seq(10, 50, 10)  # the test sample sizes
N = length(n_try)      # count them

# containers: 
out_sig = matrix(NA, I, N) # matrix with I rows and N columns
out_prc = matrix(NA, I, N) # matrix with I rows and N columns

for (n in 1:N) {
  for (i in 1:I) {
    tmp = sim_fit(n = n_try[n])     # run sim
    out_sig[i,n] = tmp["sig_pval"]  # extract and store significance metric
    out_prc[i,n] = tmp["prc_est"]   # extract and store precision metric
  }
}

par(mfrow = c(1,2))
plot(apply(out_sig, 2, mean) ~ n_try, type = "l",
     xlab = "Tagged Fish per Treatment",
     ylab = "Probability of Finding Effect (Power)")
plot(apply(out_prc, 2, mean) ~ n_try, type = "l",
     xlab = "Tagged Fish per Treatment",
     ylab = "Probability of A Precise Estimate")
```

It seems that even if you tagged `r n_try[N]` fish per treatment, you would have a `r round(mean(out_prc[,N]),2) * 100`% chance of estimating that the mortality rate is between 0.2 and 0.3 if it was truly 0.25.

You and your colleagues consider these results and determine that you will need to somehow aquire more funds to tag more fish in the small-scale study in order to a high level of confidence in the results.

### Harvest Policy Analysis {#harv-ex}

In this example, you will simulate population dynamics under a more realistic model than in Sections \@ref(for-loops) and \@ref(adv-funcs) for the purposes of evaluating different harvest policies. 

Suppose you are a fisheries research biologist, and a large commercial fishery for pink salmon (_Oncorhynchus gorbuscha_) takes place in your district. For the past 10 years, it has been fished with an exploitation rate of 40% (40% of the fish that return each year have been harvested, exploitation rate is abbreviated by $U$). The management plan is up for evaluation this year, and your supervisor has asked you to prepare an analysis that determines if more yields could be taken if a different exploitation rate were to be used in the future.

Based on historical data, your best understanding implies that the stock is driven by Ricker spawner-recruit dynamics. That is, the total number of fish that return this year (recruits) are a function of the total number of spawners that spawned the recruits this year. The functional form of the Ricker relationship can be written this way:

$$R_t = \alpha S_{t-1} e^{-\beta S_{t-1} + \varepsilon_t} ,\varepsilon_t \sim N(0,\sigma)$$

where $\alpha$ is a parameter representing the maximum recruits per spawner (obtained at very low spawner abundances) and $\beta$ is a measure of the strength of **density-dependent mortality**. Notice that the error term is in the exponent, which makes $e^{\varepsilon_t}$ lognormal.

You have estimates of the parameters:

*  $\alpha = 6$
*  $\beta = 1 \times 10^{-7}$
*  $\sigma = 0.4$

You decide that you can build a policy analysis by simulating the stock forward through time under different exploitation rates to determine if its reasonable to expect a different exploitation rate to provide more yields than what are currently being extracted.

First, write a function for your population model. Your function must:

1.  take the parameters, dimensions (number of years), and the policy variable ($U$) as input arguments
2.  simulate the population using Ricker dynamics
3.  calculate and return the sum of the yields over the number future years you simulated.

### Mark-Recapture Evaluation {#mrc-ex}

## Resampling-Based Examples {#resample-examples}

### Permutation Test {#perm-test-ex}
### The Bootstrap {#boot-test-ex}

---

## Exercise 4

In these exercise, you will be adapting the code written in this chapter to investigate slightly different questions. You should create a new R script `Ex4.R` in your working directory for these exercises so your chapter code is left unchanged. Exercise 4A is based solely on the required material and Exercises 4B - 4F are based on the example cases. You should work through each example before attempting each of the later exercises.

_The solutions to this exercise are found at the end of this book ([here](#ex4a-answers)). You are **strongly recommended** to make a good attempt at completing this exercise on your own and only look at the solutions when you are truly stumped._

### Exercise 4A: Required Material Only {-}

These questions are based on the material in Sections \@ref(randomness) - \@ref(mc-summaries) only.

1.  
2.  
3.
4.
5.
6.

[Solutions](#ex4a-answers)

### Exercise 4B: Stochastic Power Analysis {-}

These questions will require you to adapt the code written in Section \@ref(power-ex)

1.  What sample size `n` do you need to have a power of 0.8 of detecting a significant difference between the two tagging methods?
2.  How do the inferences from the power analysis change if you are interested in `p_new = 0.4` instead of `p_new = 0.25`? Do you need to tag more or fewer fish in this case?
3.  Your analysis takes a bit of time to run so you are interested in tracking its progress. Add a progress message to your nested `for` loop that will print the sample size currently being analyzed: 


```{r, eval = F}
for (n in 1:N) {
  cat("\r", "Sample Size = ", n_try[n])
  for (i in 1:I) {
    ...
  }
}
```

[Solutions](#ex4b-answers)

### Exercise 4C: Harvest Policy Analysis {-}

These questions will require you to adapt the code written in Section \@ref(harv-ex)

1.  
2.  
3.

[Solutions](#ex4c-answers)

### Exercise 4D: Mark-Recapture Evaluation {-}

These questions will require you to adapt the code written in Section \@ref(mrc-ex)

1.  
2.  
3.

[Solutions](#ex4d-answers)

### Exercise 4E: Permutation Tests {-}

These questions will require you to adapt the code written in Section \@ref(perm-test-ex)

1.  
2.  
3.

[Solutions](#ex4e-answers)

### Exercise 4F: The Bootstrap {-}

These questions will require you to adapt the code written in Section \@ref(boot-test-ex)

1.  
2.  
3.

[Solutions](#ex4f-answers)
