# Mapping and Spatial Analysis {#ch6}

This is Henry's writing. He is introducing Chapter 6 and why spatial stuff in R is easy and possibly better than using GIS.

## This is the First Section {#fsec}

Henry puts some stuff about the first section here. If Henry wants to reference some code in text, he can do `this`. If Henry wishes to execute some code and print the result in text, he can do this: `r mean(rnorm(100))`.

If Henry has a footnote, he can do this^[This is Henry's footnote].

### A subsection of the First Section

Henry has a specific subtopic under the first section.

## This is the Second Section

Henry can link to Section \@ref(fsec) like this. It won't work until it is incorporated into Ben's bookdown framework.

Henry can embed code like this:

```{r}
mean(rnorm(100))
```

Henry can hide the code but execute it and show the result like this:

```{r, echo = F}
mean(rnorm(100))
```

Henry can have this code ran but hide the result like this:

```{r, results="hide"}
mean(rnorm(100))
```

Henry can include code but not execute it like this:

```{r, eval = F}
mean(rnorm(100))
```

Henry can also include plots like this:

```{r, fig.align = "center"}
hist(rnorm(100))
```

If Henry loads a package and wishes to hide any warnings and messages, he can use this:

```{r, message = F, warning = F}
library(dplyr)
```

